---
title: "00-Data-Processing-for-Food-Banks-Accessibility"
---

## Introduction

In this notebook we create some examples of processing the data for the analysis of accessibility to food banks in Hamilton.

## Preliminaries

Clear environment:
```{r}
rm(list = ls())
```

Load packages used in the notebook:
```{r message = FALSE}
library(cancensus)
#library(disk.frame)
#library(pycno)
library(readr)
#library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
library(sf)
library(smoothr)
library(tidyverse)
library(tmap)
library(tmaptools)
```

Data used in this notebook was retrieved from three sources: Open Hamilton, Transportation Tomorrow Survey, and the 2016 Canadian Census. Open Hamilton is an online repository of data from the City of Hamilton. The provenance of the datasets is below.

## Provenance of data: 

### Census Data

First, load the water boundary for Hamilton Harbour:

```{r}
load("input-data-files/hamilton_harbour.RData")
```

### Census data

We can get an API key to use `cancensus` from [CensusMapper](https://censusmapper.ca/). Once you have an API key, it can be stored locally this way:
```{r eval=FALSE}
set_api_key("YOUR API KEY HERE")
```

Once you have your API you can use `cancensus` like in this example.

Use `cancensus` to check and download census data. Check the regions:
```{r message = FALSE}
list_census_regions('CA16') %>% 
  dplyr::filter(level == "CSD", name %in% c("Hamilton"))
```

We need PR_UID 35.

List census data sets:
```{r}
list_census_datasets()
```

We need CA16.

Search variables for DA-level data:
```{r}
find_census_vectors("age", 
                    dataset = "CA16", 
                    type = "total", 
                    query_type = "keyword", 
                    interactive = FALSE)
```

We need v_CA16_2406 to v_CA16_2425: these are the number of households at certain levels of total income. Retrieve data:
```{r}
data_da_2016 <- get_census(dataset='CA16', 
                          regions=list(CSD=c("3525005")),
                          vectors=c("v_CA16_2406",
                                    "v_CA16_2407",
                                    "v_CA16_2408",
                                    "v_CA16_2409",
                                    "v_CA16_2410",
                                    "v_CA16_2411",
                                    "v_CA16_2412",
                                    "v_CA16_2413",
                                    "v_CA16_2414",
                                    "v_CA16_2415",
                                    "v_CA16_2416",
                                    "v_CA16_2417",
                                    "v_CA16_2418",
                                    "v_CA16_2419",
                                    "v_CA16_2420",
                                    "v_CA16_2421",
                                    "v_CA16_2422",
                                    "v_CA16_2423",
                                    "v_CA16_2424",
                                    "v_CA16_2425",
                                    "v_CA16_2426",
                                    "v_CA16_2397"), # median hhinc
                          level='DA', 
                          use_cache = FALSE,
                          geo_format = "sf") %>%
  # project to NAD 1983 Zone 17N
  st_transform(crs = 26917)
```

#### Prepare Geography

Clip `data_da_2016` to remove the Hamilton Harbour from the CSD polygons:
```{r warning = FALSE}
data_da_2016 <- st_difference(data_da_2016, st_union(st_combine(hamilton_harbour %>% filter(NAME == "Hamilton Harbour")))) %>% drop_crumbs(threshold = 1)

tm_shape(data_da_2016) + 
  tm_fill(col = "Population")
```

Create a Hamilton CSD (municipal) boundary polygon:
```{r}
hamilton_boundary <- data_da_2016 %>%
  group_by(CSD_UID) %>%
  summarize(Population = sum(Population))

tm_shape(hamilton_boundary) + 
  tm_fill(col = "Population")
```

#### Prepare Summary Measures

Calculate number of households with incomes < \$40k per DA and proportion of households with incomes < \$40k per DA:
```{r}
data_da_2016 <- data_da_2016 %>%
  #filter(`Region Name` == "Hamilton") %>%
  mutate(income_less40k = (`v_CA16_2406: Under $5,000` + 
                             `v_CA16_2407: $5,000 to $9,999` + 
                             `v_CA16_2408: $10,000 to $14,999` +
                             `v_CA16_2409: $15,000 to $19,999` +
                             `v_CA16_2410: $20,000 to $24,999` +  
                             `v_CA16_2411: $25,000 to $29,999` +
                             `v_CA16_2412: $30,000 to $34,999` +
                             `v_CA16_2413: $35,000 to $39,999`),
         p_inc_less40k = income_less40k/Households)
```

Plot population:
```{r}
tm_shape(data_da_2016) + 
  tm_fill(col = "income_less40k", alpha = 0.8)
```

```{r income-quintiles, include=FALSE}
# Data preparation
## Calculate quintiles of median household income in Hamilton:
mhi_q <- data_da_2016 %>%
  dplyr::filter(`Region Name` == "Hamilton") %>%
  pull(`v_CA16_2397: Median total income of households in 2015 ($)`) %>%
  quantile(c(0, 0.2, 0.4, 0.6, 0.8, 1), 
           na.rm = TRUE)
```

Label DAs with income quintiles:
```{r}
data_da_2016 <- data_da_2016 %>%
  rename(Median_household_income = `v_CA16_2397: Median total income of households in 2015 ($)`) %>%
  mutate(income_quintile = case_when(Median_household_income < mhi_q[2] ~ "Bottom 20%",
                                    Median_household_income >= mhi_q[2] & Median_household_income < mhi_q[3] ~ "Fourth 20%",
                                    Median_household_income >= mhi_q[3] & Median_household_income < mhi_q[4] ~ "Third 20%",
                                    Median_household_income >= mhi_q[4] & Median_household_income < mhi_q[5] ~ "Second 20%",
                                    Median_household_income >= mhi_q[5] ~ "Top 20%"),
         income_quintile = factor(income_quintile,
                                levels = c("Top 20%",
                                           "Second 20%",
                                           "Third 20%",
                                           "Fourth 20%",
                                           "Bottom 20%"),
                                ordered = TRUE))
```

### Location of food banks

Read file with locations:
```{r}
foodbank_locations <- read_csv("input-data-files/foodbanks-hamilton.csv",
                               col_types = cols(ID = col_double(),
                                    NAME = col_character(),
                                    TYPE = col_character(),
                                    ADDRESS = col_character(),
                                    STATUS = col_character(),
                                    latitude = col_double(),
                                    longitude = col_double(),
                                    name = col_character(),
                                    desc = col_character(),
                                    source = col_character(),
                                    precision = col_character()
                               ))
```

`STATUS` corresponds to the food bank's status during the pandemic. Let's simplify it to a yes-no factor for whether the service remained open or closed after COVID began and then filter out inapplicable services:
```{r}
foodbank_locations <- foodbank_locations %>%
  mutate(STATUS = replace_na(STATUS, "Open"),
         COVIDClose = if_else(STATUS == "Cancelled", true = "Yes", false = "No")) %>%
  
  # filter out Meals on Wheels, Student Nutrition programs, Shopping and Transportation, & Good Food Box Programs
  filter(TYPE %in% c("Community Kitchen", "Congregate Dining", "Community Meals", "Food Banks", "Free Meals", "Low Cost Meals"))
```

Convert to simple features:
```{r message = FALSE}
foodbank_locations <- foodbank_locations %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = 26917) %>% # project to NAD 1983 Zone 17N
  
  # filter out points beyond the Hamilton CMA (there's one point in Brantford)
  filter(st_intersects(x = geometry, y = hamilton_boundary, sparse = FALSE))
```

Map location of foodbanks:
```{r}
ggplot() +
  geom_sf(data = hamilton_boundary) +
  geom_sf(data = foodbank_locations,
          aes(shape = COVIDClose,
              color = COVIDClose))
```

### Origin parcel locations

This is based on parcel-level data, approximated to the nearest point on the road network. For processing details, see:

https://github.com/paezha/Accessibility-to-Schools-Hamilton-Equity

```{r}
load("input-data-files/dwelling_network_points_2016.RData")

dwelling_network_points_2016 <- dwelling_network_points_2016 %>%
  st_transform(crs = 26917) # project to NAD 1983 Zone 17N
```

### Transportation Tomorrow Survey

We downloaded a cross-tabulation of travel mode by household income by traffic analysis zone. We also downloaded a cross-tabulation of age of person by household income by traffic analysis zone.

Read the shape file with the traffic analysis zone boundaries:
```{r}
tts_taz <- st_read("input-data-files/tts2016_2006zn_2018_region.shp") %>% 
  st_transform(crs = 26917) %>% # project to NAD 1983 Zone 17N
  st_make_valid() # fix a few topolgy errors in the boundaries
```

Filter Hamilton (region 6):
```{r}
tts_taz <- tts_taz %>%
  dplyr::filter(region == 6)

ggplot() +
  geom_sf(data = tts_taz)
```

Read transportation data:
```{r}
travel_modes <- read_csv("input-data-files/ttsCross70628.csv",
                         col_types = cols(gta06_hhld = col_double(),
                                          income = col_double(),
                                          total = col_double(),
                                          Mode = col_character()
                                          ))
```

Read population by age by income:
```{r}
age_income <- read_csv("input-data-files/ttsCross65713.csv",
                         col_types = cols(gta06_hhld = col_double(),
                                          age = col_double(),
                                          population = col_double(),
                                          income = col_double()
                                          ))
```

Code | Income
-- | --
1	| \$0 to $14999
2	| \$15000 to $39999
3	| \$40000 to $59999
4	| \$60000 to $99999
5	| \$100000 to $124999
6	| \$125000 and above
7	Decline / don't know

Filter modes:
```{r}
travel_modes <- travel_modes %>%
  dplyr::filter(Mode == "Auto driver" | 
                  Mode == "Auto passenger" |
                  Mode == "Transit (Excluding GO rail)" |
                  Mode == "Walk")
```

Filter low income:
```{r}
age_income <- age_income %>%
  dplyr::filter(income <= 2)
```

### Income less than $40,000

Filter income:
```{r}
travel_modes_1 <- travel_modes %>%
  dplyr::filter(income >= 1 & income <= 2)
```

Summarize trips by mode by traffic analysis zone:
```{r}
travel_modes_1 <- travel_modes_1 %>%
  group_by(gta06_hhld, Mode) %>%
  summarize(trips = sum(total),
            .groups = "drop")
```

Summarize population by age by income by traffic analysis zone. Children:
```{r}
low_income_children <- age_income %>%
  dplyr::filter(age <= 18) %>%
  rename(TAZUID = gta06_hhld) %>%
  group_by(TAZUID) %>%
  summarize(children = sum(population),
            .groups = "drop")
```

Adults:
```{r}
low_income_adults <- age_income %>%
  dplyr::filter(age > 18 & age < 65) %>%
  rename(TAZUID = gta06_hhld) %>%
  group_by(TAZUID) %>%
  summarize(adults = sum(population),
            .groups = "drop")
```

Seniors:
```{r}
low_income_seniors <- age_income %>%
  dplyr::filter(age >= 65) %>%
  rename(TAZUID = gta06_hhld) %>%
  group_by(TAZUID) %>%
  summarize(seniors = sum(population),
            .groups = "drop")
```

Join tables with population by age and add geometry:
```{r}
population_age_income <- tts_taz %>%
  transmute(TAZUID = gta06, 
            geometry) %>%
  left_join(low_income_children,
            by = "TAZUID") %>%
  left_join(low_income_adults,
            by = "TAZUID") %>%
  left_join(low_income_seniors,
            by = "TAZUID") %>%
  mutate(children = replace_na(children, 0),
         adults = replace_na(adults, 0),
         seniors = replace_na(seniors, 0))
```

Pivot wider and rename columns:
```{r}
travel_modes_1 <- travel_modes_1 %>%
  rename(TAZUID = gta06_hhld) %>%
  pivot_wider(names_from = Mode,
              values_from = trips,
              values_fill = 0) %>%
  mutate(Driver = `Auto driver` + `Auto passenger`) %>%
  rename(Transit = `Transit (Excluding GO rail)`) %>%
  select(-c(`Auto driver`, `Auto passenger`))
```

Join geometry:
```{r}
travel_modes_1 <- tts_taz %>%
  dplyr::transmute(TAZUID = gta06, geometry) %>%
  left_join(travel_modes_1,
            by = "TAZUID") %>%
  st_as_sf()
```

Plot trips as driver:
```{r}
ggplot() +
  geom_sf(data = travel_modes_1,
          aes(fill = Driver))
```

Plot trips transit:
```{r}
ggplot() +
  geom_sf(data = travel_modes_1,
          aes(fill = Transit))
```

Plot walking trips:
```{r}
ggplot() +
  geom_sf(data = travel_modes_1,
          aes(fill = Walk))
```

Rename table:
```{r}
modes_less40k <- travel_modes_1
```

## Before saving

Ensure that parcel ids are unique:
```{r}
dwelling_network_points_2016 <- dwelling_network_points_2016 %>% distinct(ID, .keep_all = TRUE)
```

## Save data to disk

Save data files:
```{r}
save(data_da_2016, file = "output-data-files/data_da_2016.RData", compress = "xz")
save(hamilton_boundary, file = "output-data-files/hamilton_boundary.RData", compress = "xz")
save(dwelling_network_points_2016, file = "output-data-files/dwelling_network_points_2016.RData", compress = "xz")
save(foodbank_locations, file = "output-data-files/foodbank_locations.RData", compress = "xz")
save(modes_less40k, file = "output-data-files/modes_less40k.RData", compress = "xz")
save(population_age_income, file = "output-data-files/population_age_income.RData", compress = "xz")
#save(urban_types, file = "output-data-files/urban_types.RData", compress = "xz")
```

Copy files to folder `data` of the package.
