---
title: "01-Routing-Car"
---

## Introduction

In this notebook we do the routing for the analysis of accessibility to food banks in Hamilton. This is required to obtain a travel time matrix. The routing is done using the package `r5r`.

## Preliminaries

Clear environment:
```{r}
rm(list = ls())
```

Load packages used in the notebook:
```{r message = FALSE}
library(cancensus)
library(disk.frame)
#library(pycno)
library(readr)
library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
library(sf)
library(smoothr)
library(tidyverse)
library(tmap)
library(tmaptools)
```

Define some parameters for `disk.frame` and `r5r`:
```{r}
setup_disk.frame()
options(scipen = 999)
options(java.parameters = "-Xmx6G")
options(future.globals.maxSize = Inf)
```

Data used in this notebook was retrieved from two sources: Open Hamilton and the 2016 Canadian Census. Open Hamilton is an online repository of data from the City of Hamilton. The provenance of the datasets is below.

## Provenance of data: 

### Hamilton CMA boundaries

Statistics Canada. Load:
```{r}
load("input-data-files/hamilton_cma.RData")
#load("input-data-files/hamilton_da_2016.RData")
```

### Location of food banks

Read file with food bank locations:
```{r}
load("output-data-files/foodbank_locations.RData")
```

### Location of origin parcels

This is based on parcel-level data, approximated to the nearest point on the road network. For processing details, see:

https://github.com/paezha/Accessibility-to-Schools-Hamilton-Equity

```{r}
load("input-data-files/dwelling_network_points_2016.RData")
```

## Routing for entire system

I used [BBBike](https://download.bbbike.org/osm/bbbike/) to extract OSM data for Hamilton. The name of the file is `planet_-80.279,43.048_-79.318,43.459.osm.pbf`. Copy to folder `r5_graph`.

Set Up R5 Routing. First define the path to where the graph is located:
```{r set up r5 path, include=FALSE}
r5_path <- file.path("./r5_graph")
```

Download and import GTFS (Hamilton transit data)
```{r eval=FALSE}
download.file(url = "https://transitfeeds.com/p/hamilton-street-railway/31/latest/download", destfile = file.path(r5_path, "HSR_transit.zip"), mode = "wb")
```

Download and import GTFS (Hamilton transit data)
```{r eval=FALSE}
download.file(url = "https://transitfeeds.com/p/burlington-transit/294/latest/download", destfile = file.path(r5_path, "Burlington_transit.zip"), mode = "wb")
```

Build the graph:
```{r build-graph, include = FALSE}
r5_hamilton_cma <- setup_r5(data_path = r5_path, verbose = FALSE)
```

Prepare Input Data for `r5r`. The origins are the coordinates of the parcels and the destinations the coordinates of the pharmacies:
```{r prepare-inputs}
# save origins in format expected by R5R (id, lon, lat)
origins_i <- data.frame(ID = dwelling_network_points_2016$ID, 
              dwelling_network_points_2016 %>%
                st_transform(crs = 4326) %>%
                st_coordinates()) %>%
    rename(lon = X, lat = Y, id = ID) %>%
    dplyr::select(id, lon, lat)

# now foodbanks sites
destinations_j <- data.frame(ID = foodbank_locations$ID, 
              foodbank_locations %>%
                st_transform(crs = 4326) %>%
                st_coordinates()) %>%
    rename(lon = X, lat = Y, id = ID) %>%
    dplyr::select(id, lon, lat)
```

Calculate OD Matrix for car:
```{r calculate car od matrix}
# set up batching according to how many origin rows to process at one time
chunksize = 2000 # larger chunks for walking will give enough origins in each chunk to allow multiprocessing to spin up with R5R
num_chunks = ceiling(nrow(origins_i)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(origins_i,
                          outdir = "./df/origins_i",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){ 
  origins_i_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5_hamilton_cma,
                          origins = origins_i_chunk,
                          destinations = destinations_j,
                          mode = c("CAR"),
                          mode_egress = c("WALK"),
                          departure_datetime = as.POSIXct(strptime("2021-04-05 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_walk_dist = 10000, # metres
                          max_trip_duration = 180)
  
  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "./df/output_ttm_car",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
ttm_car_time <- end.time - start.time
ttm_car_time
```

## Extract travel time matrix

```{r load od matrix for car, include=FALSE}
# connect to the walking travel time matrix disk frame
ttm_car.disk.frame <- disk.frame("./df/output_ttm_car")
#opportunities_j <- as.data.frame(nyc_cb_point) %>% select(GEOID10, total_emp) %>% rename(toId = GEOID10, o_j = total_emp)
```

Convert disk.frame to data frame:
```{r}
ttm_car <- as.data.frame(ttm_car.disk.frame) %>%
  transmute(UID = as.numeric(fromId), OBJECTID = as.numeric(toId), travel_time)
```

Make sure that the origin destination pairs are distinct in the travel time matrices:
```{r}
ttm_car <- ttm_car %>%
  distinct(UID, 
           OBJECTID,
           .keep_all = TRUE)
```

## Save data to disk

Save data files:
```{r}
#save(foodbank_locations, file = "output-data-files/foodbank_locations.RData", compress = "xz")
save(ttm_car, file = "output-data-files/ttm_car.RData", compress = "xz")
```

Copy files to folder `data` of the package.
