---
title: "02-Routing-Transit"
---

## Introduction

In this notebook we do the routing for the analysis of accessibility to food banks in Hamilton. This is required to obtain a travel time matrix. The routing is done using the package `r5r`.

## Preliminaries

Clear environment:
```{r}
rm(list = ls())
```

Load packages used in the notebook:
```{r include = FALSE}
library(cancensus)
library(disk.frame)
#library(pycno)
library(readr)
library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
library(sf)
library(smoothr)
library(tidytransit)
library(tidyverse)
library(tmap)
library(tmaptools)
```

Define some parameters for `disk.frame` and `r5r`:
```{r}
setup_disk.frame()
options(scipen = 999)
options(java.parameters = "-Xmx6G")
options(future.globals.maxSize = Inf)
```

Data used in this notebook was retrieved from two sources: Open Hamilton and the 2016 Canadian Census. Open Hamilton is an online repository of data from the City of Hamilton. The provenance of the datasets is below.

## Provenance of data: 

### Hamilton CMA boundaries

Statistics Canada. Load:
```{r}
load("input-data-files/hamilton_cma.RData")
#load("input-data-files/hamilton_da_2016.RData")
```

### Location of food banks

Read file with food bank locations:
```{r}
load("output-data-files/foodbank_locations.RData")
```

### Location of origin parcels

This is based on parcel-level data, approximated to the nearest point on the road network. For processing details, see:

https://github.com/paezha/Accessibility-to-Schools-Hamilton-Equity

```{r}
load("input-data-files/dwelling_network_points_2016.RData")
```

## Routing for entire system

I used [BBBike](https://download.bbbike.org/osm/bbbike/) to extract OSM data for Hamilton. The name of the file is `planet_-80.279,43.048_-79.318,43.459.osm.pbf`. Copy to folder `r5_graph`.

Set Up R5 Routing. First define the path to where the graph is located:
```{r set up r5 path, include=FALSE}
r5_path <- file.path("./r5_graph")
```

Download and import GTFS files

```{r eval=FALSE}
# gtfs files
download.file(url = "http://transitfeeds.com/p/hamilton-street-railway/31/20210122/download", 
              destfile = file.path(r5_path, "HSR.zip"), 
              mode = "wb")

download.file(url = "http://transitfeeds.com/p/burlington-transit/294/20210315/download", 
              destfile = file.path(r5_path, "BURT.zip"), 
              mode = "wb")
```

Build the graph:

```{r build-graph, include = FALSE}
r5_hamilton_cma <- setup_r5(data_path = r5_path, verbose = FALSE)
```

Prepare Input Data for `r5r`. The origins are the coordinates of the parcels and the destinations the coordinates of the pharmacies:
```{r prepare-inputs}
# save origins in format expected by R5R (id, lon, lat)
origins_i <- data.frame(ID = dwelling_network_points_2016$ID, 
              dwelling_network_points_2016 %>%
                st_transform(crs = 4326) %>%
                st_coordinates()) %>%
    rename(lon = X, lat = Y, id = ID) %>%
    dplyr::select(id, lon, lat)

# now foodbanks sites
destinations_j <- data.frame(ID = foodbank_locations$ID, 
              foodbank_locations %>%
                st_transform(crs = 4326) %>%
                st_coordinates()) %>%
    rename(lon = X, lat = Y, id = ID) %>%
    dplyr::select(id, lon, lat)
```

Calculate OD Matrix for transit:

```{r calculate transit od matrix}
ttm_transit <- travel_time_matrix(r5_hamilton_cma,
                          origins = origins_i,
                          destinations = destinations_j,
                          mode = c("WALK", "TRANSIT"),
                          mode_egress = c("WALK"),
                          departure_datetime = as.POSIXct(strptime("2021-03-30 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_walk_dist = 2000, # metres
                          max_trip_duration = 300, 
                          verbose = FALSE)
```

## Extract travel time matrix

```{r}
ttm_transit <- ttm_transit %>%
  transmute(UID = as.numeric(fromId), OBJECTID = as.numeric(toId), travel_time) %>%
  distinct(UID, OBJECTID, .keep_all = TRUE)
```

## Save data to disk

Save data files:
```{r}
save(ttm_transit, file = "output-data-files/ttm_transit.RData", compress = "xz")
```

Copy files to folder `data` of the package.
